Ab Initio Effective One-Electron Potential Operators: Applications for Charge-Transfer Energy in Effective Fragment Potentials

### 2019-12-09 papers/2_oep_technique.CT/paper-r3.tex
Revisions in II paper added. III paper initialized.

Four complexes: 
(i) (H$_2$O)$_2$, 
(ii) H$_2$O--CH$_3$OH, 
(iii) H$_2$O--NH$_4^+$ and 
(iv) NO$_3^-$--NH$_4^+$,
were chosen as model systems to analyse the asymptotic dependence 
of CT/HF0 energy. 
The reference (zero\hyp{}displacement) geometries
were obtained by
performing energy\hyp{}optimizations at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian~16} quantum chemistry program package.\cite{Gaussian16}
Subsequently, 30 displaced geometries for each model complex
were obtained by translating one of the monomers along the vector 
co\hyp{}linear with the H-bond or N--N distance in the case of ammonium nitrate.
The reference structures as well as the translation vectors are
indicated in the insets of Figure~\ref{f:fig-1} and S1.
To perform statistical error analysis, structural databases
of bi\hyp{}molecular complexes in the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized.
In particular, the subsets from the NCB31 database were separately considered:
the HB6/04 hydrogen bonding database,\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the DI6/04 dipole interaction database,\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the CT7/04 charge-transfer complex database,\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the WI7/05 weak interaction database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005,Zhao.Truhlar.JPCA.2005}
and 
the PPS5/05 $\pi$-$\pi$ stacking database.\cite{Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005,Zhao.Truhlar.JPCA.2005}

@@

{
\renewcommand{\arraystretch}{1.4}
\begin{table}[h]
\caption[Minimal auxiliary basis set optimized for OEP\hyp{}based CT/HF0 calculations.]
{{\bf CPU timings in miliseconds of CT/HF0 single point energy calculations
for water-methanol complex at 6-311++G(d,p) primary basis set\footnotemark[1]}
}
\label{t:5}
\begin{ruledtabular}
\begin{tabular}{llccc}
                      && Water-Methanol && Water dimer \\
\cline{3-5}
OL                    &&14749.1 &&  2909.9\\     %14716.1 &&  2876.9 added 33 sec and 17 sec for ERI/AO calculation from timer.dat
EFP2                  &&  42.3  &&    15.4\\
OEP/aug-cc-pVQZ-jkfit &&   4.7  &&     1.8\\
OEP/aug-cc-pVTZ-jkfit &&   4.0  &&     1.5\\
OEP/aug-cc-pVDZ-jkfit &&   3.8  &&     1.4\\
OEP/6-311++G(d,p)     &&   3.0  &&     1.0\\
OEP/mini\footnotemark[2]&& 2.5  &&     0.75\\
\end{tabular}
\end{ruledtabular}
%
\footnotetext[1]{1.2 GHz AMD EPYC\texttrademark{} 7301 16-Core Processor, calculations performed on 1 core. 
See also the implementation details in Section~\ref{s:4.calculations} and Section~\ref{ss:5.2.cost}.}
\footnotetext[2]{This work, Table~\ref{t:3}.}
%
\end{table}
}
%

In this work, the minimal uncontracted auxiliary basis sets were optimized for water and methanol
that consist of only $s$ and $p$\hyp{}type functions,
and the orbital exponents are shown in Table~\ref{t:3},
whereas the procedure is outlined in Appendix~\ref{a:auxiliary-basis}.
Unfortunately, we were not able to successfully fit the auxiliary basis sets
for NH$_4^+$ and NO$_3^-$ using current approach
and the basis set optimization falls out of scope of this work.
Nevertheless, we found that in the case of neutral systems minimal basis
sets are already sufficient to reach required accuracy.
In Figure~\ref{f:fig-4} the asymptotic dependence of the CT/HF0 energy
for water\hyp{}methanol system is correctly reproduced by the OEP/mini variant
which
performs comparably well as the OEP/aug-cc-pVQZ-jkfit. 
The total time required for evaluation of the CT/HF0 energies by using EFP2 and OEP
models was measured for this system and also for the water-water system in their reference geometries,
and the results are shown in Table~\ref{t:5}. 
Profiling of the code for OL, EFP2 and OEP methods was performed 
for all the computational operations required for a
single point calculation in a hypothetical sequential run on multiple
geometries. Therefore, calculations of ERI's in AO basis 
along with their four\hyp{}index transformation to MO basis were included in
time profiling of the OL routine in our computer code. For the EFP2 
and OEP methods, all the calculables were taken into account in the profiling,
i.e., calculations of OEI's in AO basis and their two\hyp{}index transformations
to MO basis. Calculations of canonical orbital 
energies and LCAO\hyp{}MO coefficients through the self\hyp{}consistent field
HF procedure, as well as calculations of the
OEP matrices from Eq.~\eqref{e:v-oep.ct} 
via EDF-2 scheme and atomic effective charges from Eq.~\eqref{e:a1-v.oep-camm.q}
using the OED in Eq.~\eqref{e:oed-group-iii}
was not included because these quantities constitute the effective fragments
by definition.
The above profiling settings allows one to compare the performances
of OL, EFP2 and OEP methods as effective fragment parameter methods.

Even for the rather large quadruple\hyp{}$\zeta$ auxiliary basis set,
the OEP\hyp{}based model is roughly 9 times more efficient than evaluating the EFP2 CT energy.
Reducing the size of auxiliary basis set up to minimal basis from Table~\ref{t:3}
scales down the cost further by a factor of $1.9$ in water-methanol system,
and $2.4$ in water dimer system, reaching in total roughly 20\hyp{}fold
reduction in CPU time as compared to the EFP2 model, and more than three orders of magnitude
as compared to the OL model. Since the size of such a minimal basis
is very small as compared to the primary basis sets recommended for constructing
the EFP2 parameters, the additional superimposition costs of the auxiliary basis
should be negligible.
It is also worth mentioning here that, while using the usual 6-311++G(d,p) basis
in a role of auxiliary basis set 
gave already accurate results, the STO-3G basis as auxiliary space resulted
in CT energies of around $-8$ kcal/mol (data not shown). Therefore, the basis set optimization
described in Appendix~\ref{a:auxiliary-basis} was necessary for the minimal basis to be applicable
in the CT calculations using OEP technique, with OEP's generated by the EDF-2 scheme
developed in Section~\ref{sss:2.3.1.GDF-2}. 
%
%It is clear that the cost of the OEP\hyp{}based equations is much lesser than the cost of EFP2 method,
%and many orders of magnitude lesser than the original theory of Murrell et al.
%
\begin{figure}[t]
\includegraphics[width=0.46\textwidth]{fig-4.pdf}
\caption{\label{f:fig-4} {\bf Asymptotic dependence of the charge transfer energy
in the CT/HF0 formulation for methanol-warer complex with minimal OEP auxiliary basis.} 
Water molecule has been translated by $R$ from the starting geometry
optimized at HF/6-31+G(d,p) level,
along the vector specified in the insets.
The total interaction energy
is also shown for comparison in light grey color in this figure.
Interaction energies were obtained at the HF/6-311++G(d,p) level of theory.
In this figure, `OEP' denotes single GDF scheme with aug-cc-pVQZ-jkfit auxiliary basis set
whereas `OEP/mini` denotes double GDF scheme with the minimal auxiliary basis
given in Table~\ref{t:3}.
} 
\end{figure}
%

@@

In this work, the one\hyp{}electron effective potential operator technique
of eliminating electron repulsion integrals in the fragment\hyp{}based
theories of intermolecular interactions
was presented.
It was shown that in general case two types of OEP's
can be defined and worked out either through the density fitting or the distributed
multipole expansion. For the first group of OEP's, the density fitting was extended
and two computational treatments of OEP's were developed: the EDF-1 scheme
for RI\hyp{}type auxiliary basis sets, as well as the EDF-2 scheme, for
essentially all other types of basis sets. 
The OEP technique was then applied to calculating the charge transfer energy
in condensed phases simulations. 
The presented validation of the OEP technique
against the Otto\hyp{}Ladik CT/HF0 model as parent theory showed that in most cases
elimination of ERI's is quantitatively accurate, with few exceptions
where only qualitative accuracy was found. 
Finally, it was concluded that, while the proposed OEP\hyp{}based Otto\hyp{}Ladik
model is of comparable accuracy as the CT/HF0 formulation within the
state\hyp{}of\hyp{}the\hyp{}art EFP2 model, it significantly outperforms the latter
in terms of computational efficiency reaching speedups up to 20 times.
It should be emphasized that there is still more room for optimization of the
OEP technique. First of all, not only group (i) terms, but also
all the summations over the virtual orbitals can be optimized,
e.g., by using the QUAMBO's, following Xu and Gordon.\cite{Xu.Gordon.JCP.2013}
In addition, effective atomic charges are most of the time very small
and could be neglected, leading to further cost reductions.
In addition, the possibility of the OEP\hyp{}based formulation of the direct EFP2 CT formulation
should be studied in the future as well.
%Since EFP2 method is based on the PT/HF0 foundations
%with fragments defined in gas phase state,
%interaction energies without the CT contribution
%in this model
%are
%conceptually incomplete.
Therefore, it is believed that the OEP\hyp{}based
model can be incorporated within the EFP2 method, enabling
the full (CT\hyp{}including) EFP2 energy calculations and 
molecular dynamics simulations in large systems. However, the current model
needs to be improved by further investigating the ionic
systems in detail and possibly revising the LOA and EDF approximations.
Nevertheless,
it is anticipated here that the OEP method of ERI elimination can be used in 
virtually any other 
\emph{ab initio} fragment\hyp{}based
approaches for condensed\hyp{}phase simulations, where ERI pose the
computational challenge when confronted with the size of the system.

>>>>

Four complexes: 
(i) (H$_2$O)$_2$, 
(ii) H$_2$O--CH$_3$OH, 
(iii) H$_2$O--NH$_4^+$ and 
(iv) NO$_3^-$--NH$_4^+$,
were chosen as model systems to analyse the asymptotic dependence 
of CT/HF0 energy. 
The reference (zero\hyp{}displacement) geometries
were obtained by
performing energy\hyp{}optimizations at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
Subsequently, 30 displaced geometries for each model complex
were obtained by translating one of the monomers along the vector 
co\hyp{}linear with the H-bond or N--N distance in the case of ammonium nitrate.
The reference structures as well as the translation vectors are
indicated in the insets of Figure~\ref{f:fig-1} and S1.
To perform statistical error analysis, structural databases
of bi\hyp{}molecular complexes in the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized.
In particular, the subsets from the NCB31 database were separately considered:
the HB6/04 hydrogen bonding database,\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the DI6/04 dipole interaction database,\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the CT7/04 charge-transfer complex database,\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the WI7/05 weak interaction database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005,Zhao.Truhlar.JPCA.2005}
and 
the PPS5/05 $\pi$-$\pi$ stacking database.\cite{Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005,Zhao.Truhlar.JPCA.2005}

@@

{
\renewcommand{\arraystretch}{1.4}
\begin{table}[h]
\caption[Minimal auxiliary basis set optimized for OEP\hyp{}based CT/HF0 calculations.]
{{\bf CPU timings in miliseconds of CT/HF0 single point energy calculations
for water-methanol complex at 6-311++G(d,p) primary basis set\footnotemark[1]}
}
\label{t:5}
\begin{ruledtabular}
\begin{tabular}{llccccc}
                      && \multicolumn{2}{c}{Water-Methanol} && \multicolumn{2}{c}{Water dimer} \\
\cline{3-4}
\cline{6-7}
OL                    &&1.48$\times 10^{4}$ 
                                &(-0.81)&& 2.91$\times 10^{3}$     
                                                    &(-0.85) \\  %14716.1 &&  2876.9 added 33 sec and 17 sec for ERI/AO calculation from timer.dat
EFP2                  &&  42.7  &(-1.27)&&   15.6   &(-1.06) \\
OEP/aug-cc-pVQZ-jkfit &&   4.36 &(-1.03)&&    1.64  &(-1.05) \\
OEP/aug-cc-pVTZ-jkfit &&   3.63 &(-1.07)&&    1.35  &(-1.09) \\
OEP/aug-cc-pVDZ-jkfit &&   3.60 &(-1.05)&&    1.26  &(-1.03) \\
OEP/6-311++G(d,p)     &&   2.74 &(-1.37)&&    0.933 &(-1.31) \\
OEP/mini\footnotemark[2]&& 2.25 &(-1.34)&&    0.623 &(-1.13) \\
\end{tabular}
\end{ruledtabular}
%
\footnotetext[1]{1.2 GHz AMD EPYC\texttrademark{} 7301 16-Core Processor, calculations performed on 1 core. 
CT/HF0 energies are given in parentheses for reference (kcal/mol).
See also the implementation details in Section~\ref{s:4.calculations} and Section~\ref{ss:5.2.cost}.}
\footnotetext[2]{This work, Table~\ref{t:3}.}
%
\end{table}
}
%

In this work, the minimal uncontracted auxiliary basis sets were optimized for water and methanol
that consist of only $s$ and $p$\hyp{}type functions,
and the orbital exponents are shown in Table~\ref{t:3}
whereas the procedure is outlined in Appendix~\ref{a:auxiliary-basis}.
Unfortunately, we were not able to successfully fit the auxiliary basis sets
for NH$_4^+$ and NO$_3^-$ using current approach
and the basis set optimization falls out of scope of this work.
Nevertheless, we found that in the case of neutral systems minimal basis
sets are already sufficient to reach required accuracy.
In Figure~\ref{f:fig-4} the asymptotic dependence of the CT/HF0 energy
for water\hyp{}methanol system is correctly reproduced by the OEP/mini variant
which
performs comparably well as the OEP/aug-cc-pVQZ-jkfit. 
The total time required for evaluation of the CT/HF0 energies by using EFP2 and OEP
models was measured for this system and also for the water\hyp{}water system in their reference geometries,
and the results are shown in Table~\ref{t:5}. 
Profiling of the code for OL, EFP2 and OEP methods was performed 
for all the computational operations required for a
single point calculation in a hypothetical sequential run on multiple
geometries. Therefore, calculations of ERI's in AO basis 
along with their four\hyp{}index transformation to MO basis were included in
time profiling of the OL routine in our computer code. For the EFP2 
and OEP methods, all the calculables were taken into account in the profiling,
i.e., calculations of OEI's in AO basis and their two\hyp{}index transformations
to MO basis. Calculations of canonical orbital 
energies and LCAO\hyp{}MO coefficients through the self\hyp{}consistent field
HF procedure, as well as calculations of the
OEP matrices from Eq.~\eqref{e:v-oep.ct} 
via EDF-2 scheme and atomic effective charges from Eq.~\eqref{e:a1-v.oep-camm.q}
using the OED in Eq.~\eqref{e:oed-group-iii}
were not included because these quantities constitute the effective fragments
by definition.
The above profiling settings allows one to compare the performances
of the OL, EFP2 and OEP methods as effective fragment parameter methods.

Even for the rather large quadruple\hyp{}$\zeta$ auxiliary basis set,
the OEP\hyp{}based model is roughly 10 times more efficient than evaluating the EFP2 CT energy.
Reducing the size of auxiliary basis set up to minimal basis from Table~\ref{t:3}
scales down the cost further by a factor of $2.0$ in water\hyp{}methanol system,
and $2.6$ in water dimer system, reaching in total roughly 20\hyp{}fold
reduction in CPU time as compared to the EFP2 model, and more than three orders of magnitude
as compared to the OL model. Since the size of such a minimal basis
is very small as compared to the primary basis sets recommended for constructing
the EFP2 parameters, the additional superimposition costs of the auxiliary basis
should be negligible. 
It is also worth mentioning here that, while using the usual 6-311++G(d,p) basis
in a role of auxiliary basis set 
gave already accurate results, the minimal STO-3G basis as auxiliary space resulted
in CT energies of $-8.4$ and $-8.8$ kcal/mol in water\hyp{}methanol and water dimer systems (data not shown),
much lower than the reference values of $-0.81$ and $-0.85$ kcal/mol, respecively. 
Therefore, the basis set optimization
described in Appendix~\ref{a:auxiliary-basis} was necessary for the minimal basis to be applicable
in the CT calculations using OEP technique, with OEP's generated by the EDF-2 scheme
developed in Section~\ref{sss:2.3.1.GDF-2}. 
%
%It is clear that the cost of the OEP\hyp{}based equations is much lesser than the cost of EFP2 method,
%and many orders of magnitude lesser than the original theory of Murrell et al.
%
\begin{figure}[t]
\includegraphics[width=0.46\textwidth]{fig-4.pdf}
\caption{\label{f:fig-4} {\bf Asymptotic dependence of the charge transfer energy
in the CT/HF0 formulation for methanol-warer complex with minimal OEP auxiliary basis.} 
Water molecule has been translated by $R$ from the starting geometry
optimized at HF/6-31+G(d,p) level,
along the vector specified in the insets.
The total interaction energy
is also shown for comparison in light grey color in this figure.
Interaction energies were obtained at the HF/6-311++G(d,p) level of theory.
In this figure, `OEP' denotes single GDF scheme with aug-cc-pVQZ-jkfit auxiliary basis set
whereas `OEP/mini` denotes double GDF scheme with the minimal auxiliary basis
given in Table~\ref{t:3}.
} 
\end{figure}
%

@@

In this work, the one\hyp{}electron effective potential operator technique
of eliminating electron repulsion integrals in the fragment\hyp{}based
theories of intermolecular interactions
was presented.
It was shown that in general case two types of OEP's
can be defined and worked out either through the density fitting or the distributed
multipole expansion. For the first group of OEP's, the density fitting was extended
and two computational treatments of OEP's were developed: the EDF-1 scheme
for RI\hyp{}type auxiliary basis sets, as well as the EDF-2 scheme, for
essentially all other types of basis sets. 
The OEP technique was then applied to calculating the charge transfer energy
in condensed phases simulations. 
The presented validation of the OEP technique
against the Otto\hyp{}Ladik CT/HF0 model as parent theory showed that in most cases
elimination of ERI's is quantitatively accurate, with few exceptions
where only qualitative accuracy was found. 
Finally, it was concluded that, while the proposed OEP\hyp{}based Otto\hyp{}Ladik
model is of comparable accuracy as the CT/HF0 formulation within the
state\hyp{}of\hyp{}the\hyp{}art EFP2 model, it significantly outperforms the latter
in terms of computational efficiency reaching speedups up to 20 times.
It should be emphasized that there is still more room for optimization of the
OEP technique. First of all, not only group (i) terms, but also
all the summations over the virtual orbitals can be optimized,
e.g., by using the QUAMBO's, following Xu and Gordon.\cite{Xu.Gordon.JCP.2013}
Secondly, effective atomic charges $q_y^{B,(nj)}$ from Eq.~\eqref{e:ct.group-iii.final.b}
are most of the time very small
and could be neglected, leading to further cost reductions (not shown here).
In addition, the possibility of the OEP\hyp{}based formulation of the direct EFP2 CT formulation
should be studied in the future as well.
%Since EFP2 method is based on the PT/HF0 foundations
%with fragments defined in gas phase state,
%interaction energies without the CT contribution
%in this model
%are
%conceptually incomplete.
Therefore, it is believed that the OEP\hyp{}based
model can be incorporated within the EFP2 method, enabling
the full (CT\hyp{}including) EFP2 energy calculations and 
molecular dynamics simulations in large systems. However, the current model
needs to be improved by further investigating the ionic
systems in detail and possibly revising the LOA and EDF approximations.
Nevertheless,
it is anticipated here that the OEP method of ERI elimination can be used in 
virtually any other 
\emph{ab initio} fragment\hyp{}based
approaches for condensed\hyp{}phase simulations, where ERI pose the
computational challenge when confronted with the size of the system.
